---
title: "Bonanza Hares: Exploratory Report"
author: "Roshni Katrak-Adefowora"
date: "11/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
#load packages
library(tidyverse)
library(here)
library(lubridate)
library(ggbeeswarm)
library(effsize)
library(kableExtra)
library(broom)
```

### Introduction

The snowshoe hare is a keystone species found in North American forests (the name "snowshoe" comes from the large size of its hind feet). This study will explore counts and sizes of juvenile snowshoe hares in Bonanza Creek Experimental Forest from 1999 to 2012 (Kielland et al. 2017). Hare weights will be compared between Bonanza Creek sites and between males and females. In addition, the relationship between hind foot length and weight will be explored.

### Data and Analyses

Dr. Knut Kielland and colleagues have collected a variety of data on snowshoe hares, all of which has been made available to the public. The data include age, sex, weight, and hind foot length for 3197 hares over a fourteen year period (1999-2012). Various exploratory visualizations were used to observe juvenile hares trapped each year and to compare male and female juvenile weights. A two-sample t-test was conducted using a significance level (α) of 0.05 to compare male and female weights and the differences between male and females was described by the Cohen's *d* effect size. A linear regression was used to explore the relationship between hind foot length and weight of juvenile hares.

```{r}
#read in hares csv
hares <- read_csv(here("data", "bonanza_hares.csv"))
```

### Exploratory findings

#### Part 1. Juvenile snowshoe hare counts
```{r}
#i) Annual juvenile hare trap counts

j_hares <- hares %>% 
  filter(age=="j") #juveniles only

j_hare_counts <- j_hares %>% 
  mutate(date=mdy(date)) %>% #convert to Date class
  mutate(year=year(date)) %>% #add year column
  count(year) #total number of hares each year

#mean and median hare trappings
j_hare_mean_med <- j_hare_counts %>% 
  summarize(mean=mean(n, na.rm=TRUE), #add mean column
            median=median(n, na.rm=TRUE)) #add median column

ggplot(data=j_hare_counts, aes(x=year, y=n))+
  geom_bar(stat="identity", fill="maroon")+ #bar graph, maroon colored bars
  labs(x="Year",
       y="Number of Juvenile Hares Trapped")+ #label x- and y-axes
  theme_light()+
  scale_x_continuous(breaks=seq(1998, 2012, 2))+ #add label every 2 ticks from 1998 to 2012
  scale_y_continuous(breaks=seq(0, 200, 25))+ #add label every 25
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.major.x = element_blank()) #remove vertical gridlines
```

**Figure 1. Total number of juvenile hare trappings each year (1999-2012).** Data: Kielland et al. (2017).


In 2002 and 2009, there were no juvenile hare trappings, and 1999 had the greatest number of hare trappings (n=`r round(j_hare_counts$n[1],1)`). The mean hare trappings from 1999 to 2012 was `r (j_hare_mean_med$mean)` and the median was `r (j_hare_mean_med$median)`. There appears to be a negative trend overall, with hare trappings decreasing from 1999 to 2012. To standardize juvenile hare populations in future analyses, it may be beneficial to look at proportions rather than actual counts (i.e., the number of trappings relative to how many traps were set out would give us a better idea of how successful the trapping was). The researchers could also attempt to be more consistent by setting up the same number of traps on the same days each year.

#### Part 2. Difference in weight between male and female snowshoe hares and across sites

```{r}
#ii) Compare juvenile hare weights by sex and site (grid)

#site names written out (to include in faceted graph for clarity)
site.name <- c("Bonanza Black Spruce", "Bonanza Mature", "Bonanza Riparian")
names(site.name) <- c("bonbs", "bonmat", "bonrip")

#table of mean weight, standard deviation, and sample size for males, females, and NA
j_hares_weight_sum <- j_hares %>% 
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm=TRUE),
            sd=sd(weight, na.rm=TRUE),
            sample_size=n())

#plot male, female, and NA weights between the 3 sites
ggplot()+
  geom_beeswarm(data=j_hares, aes(x=sex, y=weight, color=sex),
                size=1.8,
                alpha=0.4)+ #transparency
  scale_color_manual(values = c("springgreen3","royalblue3"), na.value="darkorchid3")+ #colors for each sex
  geom_point(data=j_hares_weight_sum, aes(x=sex, y=mean_weight), #add mean points
             size=2.5)+
  geom_errorbar(data=j_hares_weight_sum, #add error bars
                aes(x=sex,
                    ymax=mean_weight + sd,
                    ymin=mean_weight - sd),
                width=0.2)+
  facet_wrap(~grid,
             labeller = labeller(grid=site.name))+ #separate graph for each site and rename sites
  labs(x="Sex",
       y="Weight (g)")+ #label x- and y-axes
  theme_light()+
  theme(legend.position = "none")+ #remove legend
  scale_x_discrete(breaks=c("f", "m", NA),
                  labels=c("Female", "Male", "NA")) #rename x-axis tick labels
```

**Figure 2. Juvenile hare weights by sex and location.** The green (female), blue (male), and purple (NA; sex not identified) points indicate individual juvenile hare weight in grams. The black line represents the error bar and the black point is the mean value for each group. Each graph is a separate site. Data: Kielland et al. (2017).


The mean hare weights for males do not appear to vary greatly between the three sites, but there is large variance within the male groups; it is the same case for female hare weights. NA hare weight is between male and female weights (`r round(j_hares_weight_sum$mean_weight[3], digits=1)` ± `r round(j_hares_weight_sum$sd[3], digits=1)`). Note the large standard deviations (large spread of the data) and the sample sizes (NA, n=`r (j_hares_weight_sum$sample_size[3])`; male, n=`r (j_hares_weight_sum$sample_size[2])` and female, n=`r (j_hares_weight_sum$sample_size[1])`).

```{r, include=FALSE}
##Exploratory graphs (histograms and QQ-plots) to check assumptions

#only include male and female, no NAs
j_male_female <- j_hares %>% 
  filter(sex=="m" | sex=="f")

#Faceted histogram - look fairly normal, especially female
ggplot(data=j_male_female, aes(x=weight))+
  geom_histogram()+
  facet_wrap(~sex) #separate graphs for male and female

#QQ-plot - look fairly linear (normal), male and female potentially have outliers on upper and lower ends
ggplot(data=j_male_female, aes(sample=weight))+
  geom_qq()+
  facet_wrap(~sex) #separate graphs for male and female

#Overall: assumptions of normality hold and can go on to perform two-sample t-test.
```

**Table 1. Descriptive statistics (mean, standard deviation, and sample size) for male and female juvenile snowshoe hare weights.** Data: Kielland et al. (2017).
```{r}
#iii) juvenile weight comparison males and females

#only include male and female (no NA)
j_sum <- j_hares_weight_sum %>% 
  filter(sex=="m" | sex=="f")

#change "f" and "m" to "Female" and "Male"
j_sum$sex[j_sum$sex=="f"] <- "Female"
j_sum$sex[j_sum$sex=="m"] <- "Male"


#use kableExtra for better presentation and rename columns
j_sum_kbl <- j_sum %>% 
kable(col.names = c("Sex",
                      "Mean weight (g)", 
                      "Standard deviation (g)",
                      "Sample size (n)")) %>% 
  kable_styling(full_width = FALSE)

j_sum_kbl

#two-sample t-test
j_male_sample <- j_hares %>% 
  filter(sex == "m") %>%
  pull(weight) #turn dataframe into vector for male sample

j_female_sample <- j_hares %>% 
  filter(sex == "f") %>%
  pull(weight) #turn dataframe into vector for female sample

j_ttest <- t.test(j_male_sample, j_female_sample) #run t-test
j_ttest_out <- broom::glance(j_ttest) #metrics in dataframe

#effect size (Cohen's d)
j_effect_size <- cohen.d(j_male_sample, j_female_sample, na.rm=TRUE)
```

On average, male hare weight is greater than female hare weight (`r round(j_hares_weight_sum$mean_weight[2], digits=1)` ± `r round(j_hares_weight_sum$sd[2], digits=1)` and `r round(j_hares_weight_sum$mean_weight[1], digits=1)` ± `r round(j_hares_weight_sum$sd[1], digits=1)`, respectively; mean ± 1 standard deviation). The actual mean difference is `r round(j_hares_weight_sum$mean_weight[2], digits=1)-round(j_hares_weight_sum$mean_weight[1], digits=1)`g and the difference in means is significant (Welch’s two-sample t-test: t(`r round(j_ttest_out$parameter, digits=2)`) = `r round(j_ttest_out$statistic, digits=2)`, p = `r round(j_ttest_out$p.value, digits=4)`), however our effect size of `r round(j_effect_size$estimate,2)` is small. Because statistical hypothesis tests are so dependent on sample size, it is possible that the small p-value, which in many cases would be considered statistically significant, is due to the large sample size.

#### Part 3. Relationship between juvenile snowshoe hare weight and hind foot length
```{r}
#iv) relationship between juvenile weight and hind foot length

#linear regression
j_lm <- lm(weight ~ hindft, data=j_hares) #weight as a function of hind foot length
j_lm_tidy <- tidy(j_lm) #tidy version
j_lm_glance <- broom::glance(j_lm) #metrics at a glance

#Pearsons r
j_hares_cor <- cor.test(j_hares$hindft, j_hares$weight)
j_hares_cor_tidy <- tidy(j_hares_cor) #tidy version

#scatterplot with linear regression line of hind foot length vs weight
ggplot(data=j_hares, aes(x=hindft, y=weight))+
  geom_point()+
  labs(x="Hind foot length (mm)",
       y="Weight (g)")+ #label x- and y-axes
  theme_light()+
  geom_smooth(method = "lm", #add linear regression line
              color = "red",
              size = 0.5,
              fill = "gray60",
              alpha = 0.4) +
  ggpubr::stat_regline_equation(label.x = 65, label.y = 1500) #add line equation
```

**Figure 1. Relationship between hind foot length (mm) and weight (g) of juvenile snowshoe hares.** Points are individual hare measurements. Linear model summary: $\beta$~1~ = `r round(j_lm_tidy$estimate[2], digits=2)`g/mm, p < 0.001, R^2^ = `r round(j_lm_glance$r.squared, digits=2)`, Pearson's *r* = `r round(j_hares_cor_tidy$estimate, digits=2)`). Data: Kielland et al. (2017).


The positive slope represents a positive correlation. The R^2^ value indicates that `r round(j_lm_glance$r.squared, digits=2)*100`% of the variation in weight (dependent variable) is determined by hind foot length (independent variable). The Pearson's *r* value indicates medium strength positive correlation for the relationship between hind foot length and weight. One assumption of linear regression that may be a concern is homoscedasticity, because looking at the graph of residuals about the fitted value, the spread does not look constant, indicating heteroscedasticity. 

## Summary

Our juvenile hares exploration revealed these findings:

- The highest number of juvenile hares captured was in 1999 and the number captured has since decreased.

- Male juvenile hares weigh more than female juvenile hares (significant p-value, small effect size)

- Hind foot length appears linearly related with weight. The relationship (by simple linear regression) is significant, with a slope of `r round(j_lm_tidy$estimate[2], 2)`g/mm.

## Citations

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14